---
title: "Tornado and Importance Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tornadoVignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 5
)
has_survival <- requireNamespace("survival", quietly = TRUE)
has_caret <- requireNamespace("caret", quietly = TRUE) & 
  requireNamespace("randomForest", quietly = TRUE)
has_glmnet <- requireNamespace("glmnet", quietly = TRUE)
require(tornado)
require(ggplot2)
```

## Overview

A tornado plot is a visualization of the range of outputs expected
from a variety of inputs, or alternatively, the sensitivity of the output to the
range of inputs.  Tornado plots have a number of features:

- The center of the tornado is plotted at the response
expected from the mean of each input variable.  
- For a given variable, the width
of the tornado is determined by the response to a change in each input variable
while holding all others constant at their mean.  Various criteria are used 
depending on the specifics of the analysis.  Three
primary criteria are implemented in this package:
    - The range of the variable.  The input is set at the maximum, minimum, and
    mean.
    - A multiplicative factor of the variable mean.  The input is set 
    at $100(1 - \alpha)$% and $100(1+\alpha)$% of the mean.
    - A quantile of the variable's distribution in the input data.  The input is set
    at $100(1 - \alpha)$ percentile and $100(1+\alpha)$ percentile 
- Variables are ordered vertically
with the widest bar at the top and narrowest at the bottom.  
- Only one variable is moved from its mean value at a time.  
- Factors or categorical variables have also
been added to these plots by plotting dots at the resulting output as each
factor is varied through all of its levels.  
- The base factor level is chosen as the input variable for the center of the 
tornado when factors are present.

An importance plot attempts to dispay the relative impact of each variable
on the model fit.  Traditionally, for linear models, the concept of importance
was expressed as the percentage of total response variable variance that is
explained by each variable, either alone, or in the presence of the other variables
in the model.  Each model type can have different measures of importance.  In
this package, the following methods are used:

- Linear Models:  percent of the total variance explained by each variable alone
and cumulatively with other variables in the model.
- Generalized Linear Models:  Same as linear models, but using the deviance instead
of the variance.
- Survival Models, LASSO, Ridge Regression:  The decrease in model Mean Squared
Error as each variable is permuted in the data set.  Permuting a varible in a 
data set remove the explanatory power of that variable relative to the 
other vriables in the model still correlated with the response variable.
- `caret` package models:  The definition of variable importance for each
model in the `caret` package is taken from the `varImp` method.  See the `caret`
package for specifics.

## Data Sets

A few standard data sets are used in these examples:

- `mtcars`:  Motor Trend Car Road Tests - predict Miles Per Gallon (mpg) based on
other car factors
- `mtcars_w_factors`: the `mtcars` dataset with a few numeric variables replaced
with categorical or factor variables where appropriate.  e.g. Automatic vs Manual
transmission
- `survival::ovarian`: Ovarian Cancer Survival Data - predict survival time
- `survival::bladder`: Bladder Cancer recurrences - risk of bladder cancer recurrence
- `iris`: Edgar Anderson's Iris Data - predict the type of iris from measurements

```{r datasets, echo=TRUE}
mtcars_w_factors <- mtcars
# Automatic vs Manual
mtcars_w_factors$am <- factor(mtcars$am)
# V or Straight cylinder arrangement
mtcars_w_factors$vs <- factor(mtcars$vs)
# number of cylinders
mtcars_w_factors$cyl <- factor(mtcars$cyl)
# number of forward gears
mtcars_w_factors$gear <- factor(mtcars$gear)
# number of carburetors
mtcars_w_factors$carb <- factor(mtcars$carb)
```

## Tornado Plot Examples

### LM

#### type = "PercentChange"

```{r lm1, echo=TRUE}
lm1 <- lm(mpg ~ cyl*wt*hp, data = mtcars)
torn1 <- tornado::tornado(lm1, type = "PercentChange", alpha = 0.10)
plot(torn1, xlabel = "MPG", geom_bar_control = list(width = 0.4))
```

#### type = "rangess"

```{r lm2, echo=TRUE}
torn2 <- tornado::tornado(lm1, type = "ranges")
plot(torn2, xlabel = "MPG", geom_bar_control = list(width = 0.4))
```

#### type = "percentiles"

```{r lm3, echo=TRUE}
torn3 <- tornado::tornado(lm1, type = "percentiles", alpha = 0.05)
plot(torn3, xlabel = "MPG", geom_bar_control = list(width = 0.4))
```

#### Including Categorical Variables

```{r lm4, echo=TRUE}
lm4 <- lm(mpg ~ cyl + wt + hp + vs, data = mtcars_w_factors)
torn4 <- tornado::tornado(lm4, type = "percentiles", alpha = 0.05)
plot(torn4, xlabel = "MPG", geom_bar_control = list(width = 0.4))
```

#### Changing Variable Names

```{r lm5, echo=TRUE}
dict <- list(old = c("cyl", "wt", "hp", "vs"),
             new = c("Cylinders", "Weight", "Horsepower", "V_or_Straight"))
torn5 <- tornado::tornado(lm4, type = "percentiles", alpha = 0.05, dict = dict)
plot(torn5, xlabel = "MPG", geom_bar_control = list(width = 0.4))
```

#### Plotting Options

```{r lm6, echo=TRUE}
dict <- list(old = c("cyl", "wt", "hp", "vs"),
             new = c("Cylinders", "Weight", "Horsepower", "V_or_Straight"))
torn5 <- tornado::tornado(lm4, type = "percentiles", alpha = 0.05)
plot(torn5, xlabel = "MPG", geom_bar_control = list(width = 0.4),
     sensitivity_colors = c("#FC8D62", "#66C2A5"),
     geom_point_control = list(size = 3, fill = "purple", col = "purple"))
```

#### Extending the ggplot

Notes:

- The plot center is at 0 and the x-axis is relabeled.  So Adding lines
to the plot should use a zero centered coordinate system
- Also, the plot is drawn using coord_flip, so the horizonal line below
is draw vertically

```{r lm7, echo=TRUE}
g <- plot(torn5, plot = FALSE, xlabel = "MPG", geom_bar_control = list(width = 0.4),
          sensitivity_colors = c("#FC8D62", "#66C2A5"),
          geom_point_control = list(size = 3, fill = "purple", col = "purple"))
g <- g + ggtitle("Test Plot")
g <- g + geom_hline(yintercept = 0, col = "black", lwd = 2)
plot(g)
```

### GLM

Predict if an engine is a V or Straight given covariates

```{r glm1, echo = TRUE}
glm1 <- glm(vs ~ wt + disp + cyl, data = mtcars, family = binomial(link = "logit"))
torn1 <- tornado::tornado(glm1, type = "ranges", alpha = 0.10)
plot(torn1, xlabel = "V or Straight Engine", geom_bar_control = list(width = 0.4))
```

### Censored Data

#### Surival Regression or Accelerated Failure

```{r censored1, echo=TRUE}
survreg1 <- survival::survreg(survival::Surv(futime, fustat) ~ ecog.ps + rx + age + resid.ds, 
                              survival::ovarian, dist = 'weibull', scale = 1)
torn1 <- tornado::tornado(survreg1, modeldata = survival::ovarian, 
                          type = "PercentChange", alpha = 0.10)
plot(torn1, xlabel = "Survival Time", geom_bar_control = list(width = 0.4))
```

#### Cox Proportional Hazards

```{r censored2, echo=TRUE}
coxph1 <- survival::coxph(survival::Surv(stop, event) ~ rx + size + number,
                          survival::bladder)
torn1 <- tornado::tornado(coxph1, modeldata = survival::bladder, type = "PercentChange",
                          alpha = 0.10)
plot(torn1, xlabel = "Risk", geom_bar_control = list(width = 0.4))
```

### Ridge Regression and LASSO

```{r glmnet1, echo=TRUE}
if (has_glmnet)
{
  form <- formula(mpg ~ cyl*wt*hp)
  mf <- model.frame(form, data=mtcars)
  mm <- model.matrix(form, mf)
  gtest <- glmnet::cv.glmnet(x = mm, y = mtcars$mpg, family = "gaussian")
  torn <- tornado::tornado(gtest, modeldata = mtcars, 
                           form = formula(mpg ~ cyl*wt*hp), 
                           s = "lambda.1se",
                           type = "PercentChange", alpha = 0.10)
  plot(torn, xlabel = "MPG", geom_bar_control = list(width = 0.4))
} else
{
  print("glmnet is not available for vignette rendering")
}
```

### Machine Learning Models

#### Regression

```{r train_regression, echo = TRUE}
if (has_caret)
{
  gtest <- caret::train(x = subset(mtcars_w_factors, select = -mpg), 
                        y = mtcars_w_factors$mpg, method = "rf")
  torn <- tornado::tornado(gtest, type = "percentiles", alpha = 0.10)
  plot(torn, xlabel = "MPG")
} else
{
  print("caret is not available for vignette rendering")
}
```

#### Classification

The `plot` method can also return a ggplot object

```{r train_section, echo = TRUE}
if (has_caret)
{
  gtest <- caret::train(x = subset(iris, select = -Species), 
                        y = iris$Species, method = "rf")
  torn <- tornado::tornado(gtest, type = "percentiles", alpha = 0.10, class_number = 1)
  g <- plot(torn, plot = FALSE, xlabel = "Probability of the Setosa Species")
  g <- g + ggtitle("Classifier caret::train 'rf', 10th to 90th perc. of each var.")
  plot(g)

  torn <- tornado::tornado(gtest, type = "percentiles", alpha = 0.10, class_number = 2)
  g <- plot(torn, plot = FALSE, xlabel = "Probability of the versicolor Species")
  plot(g)
} else
{
  print("caret is not available for vignette rendering")
}
```

## Importance Plots

### Linear Models

```{r imp.lm, echo=TRUE}
gtest <- lm(mpg ~ cyl*wt*hp + gear + carb, data = mtcars)
gtestreduced <- lm(mpg ~ 1, data = mtcars)
imp <- tornado::importance(gtest, gtestreduced)
plot(imp)
```

#### Using a dictionary to translate the variable names and chaning colors

```{r imp.lm.dict, echo=TRUE}
dict <- list(old = c("cyl", "wt", "hp", "vs", "gear", "carb"),
             new = c("Cylinders", "Weight", "Horsepower", "V_or_Straight", "Num Gears", "Num Carbs"))
imp <- tornado::importance(gtest, gtestreduced, dict = dict)
plot(imp, col_importance_alone = "#8DD3C7", 
     col_importance_cumulative = "#FFFFB3")
```

### Generalized Linear Models

```{r imp.glm, echo=TRUE}
gtest <- glm(vs ~ wt + disp + gear, data = mtcars, family = binomial(link = "logit"))
gtestreduced <- glm(vs ~ 1, data = mtcars, family = binomial(link = "logit"))
imp <- tornado::importance(gtest, gtestreduced)
plot(imp)
```

### Censored Data

```{r imp_survreg, echo=TRUE}
model_final <- survival::survreg(survival::Surv(futime, fustat) ~ ecog.ps*rx + age,
                                 data = survival::ovarian,
                                 dist = "weibull")
imp <- tornado::importance(model_final, survival::ovarian, nperm = 100)
plot(imp, geom_bar_control = list(width = 0.4, fill = "blue"))
```

### Machine Learning

#### Regression

The number of variables plotted can also be controlled

```{r train.regression.imp, echo = TRUE}
if (has_caret)
{
  gtest <- caret::train(x = subset(mtcars_w_factors, select = -mpg), 
                        y = mtcars_w_factors$mpg, method = "rf")
  imp <- tornado::importance(gtest)
  plot(imp, nvar = 7)
} else
{
  print("caret is not available for vignette rendering")
}
```

#### Classification

The `plot` method can also return a ggplot object

```{r train.classification.imp, echo = TRUE}
if (has_caret)
{
  gtest <- caret::train(x = subset(iris, select = -Species), 
                        y = iris$Species, method = "rf")
  imp <- tornado::importance(gtest)
  g <- plot(imp, plot = FALSE)
  g <- g + ggtitle("Classifier caret::train randomforest: variable importance")
  plot(g)
} else
{
  print("caret is not available for vignette rendering")
}
```
